---
title: "PA1"
author: "mutecypher"
date: "July 3, 2016"
output: html_document
---
This is the data analysis portion of the project 2 for Reproducible Results
```{r, knitr::opts_chunk$set(fig.width=12, fig.height=6, fig.path='figure/', warning=FALSE, message=FALSE)}
##create a directory if one does not exist
##install.packages("knitr", "rmarkdown)
##library(knitr)##
install.packages("ggplot2", repos="http://cran.rstudio.com/")
library(ggplot2)
fileUrl <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "data.zip", method = "curl")
data <- unzip("data.zip")
unlink(fileUrl)
data <- read.csv("./activity.csv")
ThisRanAt <- date()
```

The data was downloaded and read at `r ThisRanAt`.

Create a version of the data without NA's and aggregate the step data by date
```{r processit}
## make a version of the data without the NA's then get mean and median
data$date <- as.Date(data$date, "%Y-%m-%d")
noNAdata <- data[complete.cases(data),]
names(noNAdata) <- c("steps", "date", "interval")
## aggregate the data into a table with the steps summed by day
aggysteps <- aggregate(noNAdata$steps ~ noNAdata$date, noNAdata ,FUN = sum)
names(aggysteps) <- c("date", "steps")
write.csv(aggysteps, file = "aggysteps.csv")
```

Now make a nice histogram of the total steps taken per day

```{r histymeanmed, fig.height = 5}
png(filename = "firsthisty.png", width = 480, height = 480)
hist(aggysteps$steps)
dev.off()
meansteps <- mean(aggysteps$steps)
mediansteps <- median(aggysteps$steps)
```

The mean number of steps take per day is `r meansteps` and the median taken per day is `r mediansteps`.

```{r intervals}
install.packages("lubridate", repos="http://cran.rstudio.com/")
library(lubridate)
intervals <- aggregate(noNAdata$steps ~ noNAdata$interval, noNAdata ,FUN = mean)
## make the plot using the row names as the x value, each 5 minute interval is numbered by row
names(intervals) <- c("interval", "steps")
plot(rownames(intervals), intervals$steps, type = "l", xlab = "5 minute intervals(not times)", ylab = "average steps", main = "Average Number of Steps per Interval Over All Days")
topstep <- intervals[which.max(intervals$steps),]
names(topstep)
topstep$interval <- parse_date_time(topstep$interval, "%R")
toptime <- paste(hour(topstep$interval), minute(topstep$interval), sep = ":")
```

The most steppy time is `r toptime`.
```{r newdata}
countall <- dim(data)
withoutna <- dim(noNAdata)
difference <- countall[1] - withoutna[1]
```

There were `r difference` NAs in the data.
```{r insavgs}
imputeddata <- data
names(imputeddata) <- c("steps", "date", "interval")
## replace the NAs with the average for the interval
repintervals <- intervals
## create a vector the same dimension as the data for substitution
repintervals <- repintervals[rep(rownames(repintervals),61),]
imputeddata$steps <- replace(imputeddata$steps, which(is.na(imputeddata$steps)),repintervals[which(is.na(imputeddata$steps)),2])
## turn the steps column to numeric data,some were characters
imputeddata$steps <- as.numeric(as.character(imputeddata$steps))
## now sum the data for each day
imputedsteps <- aggregate(imputeddata$steps ~ imputeddata$date, imputeddata ,FUN = sum)
names(imputedsteps) <- c("date", "steps")
#now find the mean for steps per day
imputedmean <- mean(imputedsteps$steps)
imputedmed <- median(imputedsteps$steps)
```
The mean number of steps when the interval means were substituted for NA is: `r imputedmean`.

The meadian number of steps when the interval means were substituted for NA is: `r imputedmed`.


Now make a nice histogram of the total steps take per day

```{r histyimputed, fig.height = 5}
par(mar = c(5,4,1,1))
hist(imputedsteps$steps,xlab="Daily sum of steps", ylab =  "Days", main = "Sum of Steps taken per day with interval means substituted for NAs")
```
Per the directions, I will use the data with the filled in values for NA - the mean for each interval. 

```{r weekdays versus weekends}
install.packages("timeDate", repos = "http://cran.rstudio.com/")
library(timeDate)
weekdazedata <- imputeddata
##now add days of the week as a factor
weekdazedata$wkday <- isWeekday(weekdazedata$date)
weekdazedata$wknday <- isWeekend(weekdazedata$date)
```
Make the plot after aggregating the data

```{r getting weekend and weekdays}
weekdayonly <- subset(weekdazedata, weekdazedata$wkday == TRUE)
weekendonly <- subset(weekdazedata, weekdazedata$wknday == TRUE)
meanweekday <- aggregate(weekdayonly$steps ~ weekdayonly$interval, weekdayonly, FUN = mean)
meanweekend <- aggregate(weekendonly$steps ~ weekendonly$interval, weekendonly, FUN = mean)
names(meanweekday) <- c("interval", "steps")
names(meanweekend) <- c("interval", "steps")
par(mfrow = c(2,1))
plot(meanweekday$interval, meanweekday$steps, type = "l", xlab = "5 minute intervals(not times)", ylab = "average steps", main = "Average Number of Steps per Interval Over Weekdays")
plot(meanweekend$interval, meanweekend$steps, type = "l", xlab = "5 minute intervals(not times)", ylab = "average steps", main = "Average Number of Steps per Interval Over Weekend")